---
title: "OOPs Analysis"
output: bookdown::html_document2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_data, include = FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)
library(hexbin)
library(gridExtra)
library(bookdown)



rm(list=ls())

# Load data
df <- read.csv2("LSMScompilation_tidy.csv")
df_nontidy <- read.csv2("LSMScompilation_nontidy.csv")
tab_recall <- read.csv2("LSMScompilation_recall_nitems.csv")
```

# Search Strategy

Our search started at the World Bank’s Microdata Library, where we looked for micro-surveys from the 
Living Standards Measurement Surveys (LSMS) program. We collected data from 80 surveys from the program wich were publicly available.

```{r filter_surveys, include=FALSE}

# get list of surveys with both modules
surveys_bothmodules <- tab_recall %>% 
  select(survey) %>% 
  group_by(survey) %>% 
  mutate(count = n()) %>% 
  filter(count>1) %>% 
  select(survey) %>% 
  distinct()

# Keep only observations from selected surveys in df
df_bothm <- df %>% # df with surveys from both modules
  inner_join(surveys_bothmodules)

# keep only observations from selected surveys - nontidy
df_bothm_nontidy <- df_nontidy %>% 
  inner_join(surveys_bothmodules)
```

The datasets were further filtered to keep only those that measure OOPs twice in the household, once in the consumption module, and the health module. Out of the 80 initial surveys `r length(surveys_bothmodules$survey)` measured OOPs in both modules, which will be the sample we will use in what follows.

# Household OOPs

With two different OOP estimates for each household for `r length(surveys_bothmodules$survey)` surveys, we have the opportunity to explore how much the reported OOPs differ by module in the same household. Figure 1 shows scatterplots for each survey, where the x-axis corresponds to the OOPs in the health module, and the y-axis to the OOPs in the consumption module. The color of the hexagons in the figure represent the concentration of observations around the area, with more intense tones representing a higher concentration.

It is interesting to see that some surveys do a very good job at eliciting similar reported OOPs between modules, such as the surveys from Uganda, and Bulgaria. On the other hand, the other surveys have a clear pattern with households reporting more OOPs in the health module, with the exception of Ghana 2017.

```{r scatter_health_consumption, echo=FALSE, warning=FALSE, message=FALSE}

scatter_xy <- function(survey_name){

# Filter survey and take outliers out
data <- df_bothm_nontidy %>% 
  filter(survey==survey_name) %>%
  select(hhid_compilation,health_consumption,healthm_oops) %>% 
  na.omit() %>% 
  mutate(z_health = healthm_oops / sd(healthm_oops, na.rm = TRUE),
         z_consumption = health_consumption / sd(health_consumption, na.rm = TRUE)) %>% 
  filter(z_health<2, z_consumption<2) %>%  # take out all those with z>2 for viz purposes
  rename(`Health Module` = healthm_oops,
         `Consumption Module` = health_consumption)

# Limit for both axis for viz  
max_scale <- data %$% 
    max( max(`Health Module`, na.rm = TRUE), max(`Consumption Module`)     , na.rm = TRUE)

# Number of bins (optional to use)
nbins <- data %>% 
  nrow() / 100
  
# one special for Ghana 2013
  if (survey_name=="Ghana_2013"){
    graph <- data %>% 
      ggplot(aes(`Consumption Module`,`Health Module`)) +
      #geom_point(alpha = 1/10) +
      geom_hex(bins = 68) +
      scale_fill_gradient(low = "yellow", high = "red", limits = c(5, 200) , guide = FALSE) + #option to show the
      geom_abline(intercept = 0, slope = 1, alpha=10, linetype = "dashed", color="black") +
      scale_y_continuous(limits = c(0,max_scale)) +
      scale_x_continuous(limits = c(0,max_scale)) +
      ggtitle(survey_name) +
      theme(axis.text = element_text(size = 7), 
            axis.title = element_text(size = 7),
            plot.title = element_text(size = 11, face = "bold", hjust=0.5))
    
    graph

# All the rest  
  } else{
  
    graph <- data %>% 
    ggplot(aes(`Consumption Module`,`Health Module`)) +
    #geom_point(alpha = 1/10) +
    geom_hex() + #bins= nbins
    scale_fill_gradient(low = "yellow", high = "red", limits = c(5, 200) , guide = FALSE) + #option to show the
    geom_abline(intercept = 0, slope = 1, alpha=10, linetype = "dashed", color="black") +
    scale_y_continuous(limits = c(0,max_scale)) +
    scale_x_continuous(limits = c(0,max_scale)) +
    ggtitle(survey_name) +
    theme(axis.text = element_text(size = 7), 
            axis.title = element_text(size = 7),
            plot.title = element_text(size = 11, face = "bold", hjust=0.5))
  
  graph
  }

}

# List of surveys
bothm_surveylist <- surveys_bothmodules$survey %>% 
  as.character()

# get all the graphs
hex_graphs <- bothm_surveylist %>%  map(scatter_xy)

# Put all graphs in one page
grid.arrange(hex_graphs[[1]],hex_graphs[[2]],hex_graphs[[3]],hex_graphs[[4]],
             hex_graphs[[5]],hex_graphs[[6]],hex_graphs[[7]],hex_graphs[[8]],
             hex_graphs[[9]],hex_graphs[[10]],hex_graphs[[11]],hex_graphs[[12]],
             hex_graphs[[13]],hex_graphs[[14]],hex_graphs[[15]],hex_graphs[[16]], hex_graphs[[17]])


```

What are the characteristics of the surveys who have OOPs that better allign between their health and consumption modules. Table 1 sorts in ascending order the coefficient for each survey, obtained by a regression between the OOPs in the health module as the dependent variable, and the OOPs in the consumption module as the independent variable. The closer the coefficient is to one, the more these two variables move in tandem.

Not surprisingly, three surveys from Uganda each with only one item in the health and consumption module, have coefficients very close to one. Households reacted 


```{r simple_model_OOPs, echo=FALSE, warning=FALSE, message=FALSE}

# Create simple regression
survey_model <- function(survey_name){
  df_aux <- df_bothm_nontidy %>% filter(survey==survey_name)
  model <- lm(healthm_oops ~ health_consumption , data = df_aux)
  coef(model)[[2]]
}


coefs <- bothm_surveylist %>% map(survey_model) %>% unlist()

tab_reg <- tibble(coef = coefs, survey = bothm_surveylist) %>%
  inner_join(tab_recall) %>% pivot_wider(names_from = module, values_from = c(recall,nitems)) %>% 
  group_by(survey) %>% 
  transmute(
            `Regression Coefficient` = round(coef,2),
            `Items Health Module` = first(nitems_Health),
            `Items Consumption Module` = last(nitems_Consumption),
            `Avg. Recall Health Module` = first(recall_Health),
            `Avg. Recall Consumption Module` = round(last(recall_Consumption),1 ) ) %>% 
  rename(Survey = survey) %>% 
  distinct() %>% 
  arrange(`Regression Coefficient`)


kable(tab_reg, caption = "Surveys' similitude between OOPs and instrument characteristics") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
   footnote(general = "The number of items correspond to the number of instruments utilized in the survey to arrive at the OOPs estimate in each module. The Avg. Recall period is in weeks, and represents how far back in time households were asked to recall their expenditures")
  
```



# Average OOPs

```{r get_means , include=FALSE}

# Grab means by survey and module
tab_mean_oops <- df_bothm %>% 
  group_by(survey,module) %>% 
  summarise(mean_oops = mean(oops, na.rm = TRUE),
            median_oops = median(oops, na.rm=TRUE),
            obs = n())

# Grab OOPs of health module over OOPs of consumption module
tab_health_over_consumption <- tab_mean_oops %>% 
  group_by(survey) %>% 
  mutate(healthc_healthm = first(mean_oops) / last(mean_oops),
         mhealthc_mhealthm = first(median_oops) / last(median_oops),
         trim_mhealthc_mhealthm = if_else(mhealthc_mhealthm>2,2,mhealthc_mhealthm)) %>% 
  ungroup() %>% 
  select(survey, mhealthc_mhealthm, healthc_healthm, trim_mhealthc_mhealthm) %>% 
  distinct() %>% 
  mutate( order = seq_along(survey),
          survey2 = fct_reorder(survey, desc(order) ))

# get number where health module is higher
nhealth_higher <- tab_health_over_consumption %>% 
  filter(healthc_healthm<1) %>% 
  nrow()

```


In our selected sample of surveys the mean of OOPs in the health module was higher than in the consumption module for most surveys. As can be seen in Figure 1, where `r nhealth_higher` out of `r length(surveys_bothmodules$survey)` surveys have a share less than one, indicating that the mean OOPs in their consumption module are lower than the mean OOPs from their health module.

```{r mean_graph, echo=FALSE}
# Graph Average Means by survey
tab_health_over_consumption %>% 
  ggplot(aes( survey2 ,healthc_healthm)) +
  geom_point(color="blue") +
  coord_flip() +
  ggtitle("Mean OOPs from the consumption module over the health module") +
  xlab("") +
  ylab("Share") +
  geom_hline(yintercept=1, alpha=3, linetype = "dashed") +
  scale_y_continuous(breaks = seq(0,2,0.25), limits = c(0,2))
```

```{r prep_zero_oops, include=FALSE}

# Percentage by survey with zero oops
per_zero <- df_bothm %>% 
  group_by(survey,module) %>% 
  mutate(zero_oops = oops==0) %>% 
  summarise(mean_zero = mean(zero_oops, na.rm = TRUE)) %>% 
  ungroup() %>% 
  group_by(module) %>% 
  summarise(mean_zero2 = mean(mean_zero, na.rm = TRUE))

# Grab means for oops>0
tab_mean_oops_zero <- df_bothm %>% 
  filter(oops>0) %>% 
  group_by(survey,module) %>% 
  summarise(mean_oops = mean(oops, na.rm = TRUE),
            median_oops = median(oops, na.rm=TRUE),
            obs = n())

# Grab OOPs of health module over OOPs of consumption module
tab_health_over_consumption_zero <- tab_mean_oops_zero %>% 
  group_by(survey) %>% 
  mutate(healthc_healthm = first(mean_oops) / last(mean_oops),
         mhealthc_mhealthm = first(median_oops) / last(median_oops),
         trim_mhealthc_mhealthm = if_else(mhealthc_mhealthm>2,2,mhealthc_mhealthm)) %>% 
  ungroup() %>% 
  select(survey, mhealthc_mhealthm, healthc_healthm, trim_mhealthc_mhealthm) %>% 
  distinct() %>% 
  mutate( order = seq_along(survey),
          survey2 = fct_reorder(survey, desc(order) ))

# get number where health module is higher
nhealth_higher_zero <- tab_health_over_consumption_zero %>% 
  filter(healthc_healthm<1) %>% 
  nrow()
```


On average `r round(per_zero$mean_zero2[[1]]*100,1)`% and `r round(per_zero$mean_zero2[[2]]*100,1)`% of the households report zero OOPs in the consumption and health module respectively. Given the high percentage of households reporting zero OOPs, can we expect the mean OOPs between modules to allign better for households reporting positive OOPs? 

Figure 2 exhibits that `r nhealth_higher_zero` out of `r length(surveys_bothmodules$survey)` have higher mean OOPs in the health module than in the consumption module after excluding observations reporting zero OOPs.

```{r mean_oops_zeroplus, echo=FALSE}
# Graph Average Means by survey - positive oops
tab_health_over_consumption_zero %>% 
  ggplot(aes( survey2 ,healthc_healthm)) +
  geom_point(color="blue") +
  coord_flip() +
  ggtitle("Mean OOPs from the consumption module over the health module") +
  labs(subtitle = "For households reporting positive OOPs") +
  xlab("") +
  ylab("Share") +
  geom_hline(yintercept=1, alpha=3, linetype = "dashed") +
  scale_y_continuous(breaks = seq(0,2,0.25), limits = c(0,2))
```

```{r repetition_zeros, include=FALSE}

df_zero <- df_bothm %>% 
  group_by(hhid_compilation) %>% 
  mutate(
    c0_h0 = (first(oops)==0 & last(oops)==0),
    c1_h0 = (first(oops)>0 & last(oops)==0),
    c0_h1 = (first(oops)==0 & last(oops)>0),
    c1_h1 = (first(oops)>0 & last(oops)>0 ) )

df_zero2 <- df_zero %>% 
  select(hhid_compilation,survey, c0_h0, c1_h0, c0_h1, c1_h1) %>% 
  distinct()

# Table with concordance categories by survey
tab_zero_reps <- df_zero2 %>% 
  group_by(survey) %>% 
  summarise(both_zero = mean(c0_h0, na.rm = TRUE),
            posc_zeroh = mean(c1_h0, na.rm = TRUE),
            zeroc_posh = mean(c0_h1, na.rm = TRUE),
            posc_poch = mean(c1_h1, na.rm = TRUE))

# Calculate Average concordance across surveys
per_concordance <- tab_zero_reps %>% 
  summarise(both_zero2 = mean(both_zero, na.rm = TRUE),
            both_pos2 = mean(posc_poch, na.rm = TRUE))

per_concordance

```

# Zero OOPs

Also if households reported zero OOPs in one module, how do they respond in the next one? Across surveys households 
`r round(per_concordance$both_zero2[[1]]*100,1)`% of households report zero OOPs in both modules, and `r round(per_concordance$both_pos2[[1]]*100,1)`% report positive OOPs in both. In total `r (  round(   (per_concordance$both_zero2[[1]] + per_concordance$both_pos2[[1]])*100    ,1 )    )`% of households report concordant OOPs between modules, leaving `r 100- (  round(   (per_concordance$both_zero2[[1]] + per_concordance$both_pos2[[1]])*100    ,1 )    )`% of households that differ on their answers between modules, which is problematic from a measuring perspective.

Table 1 exhibits the concordance of household's answers between the health module and consumption module by survey. The percentages vary widely by surveys, Ghana_1991 and Ghana_2005 both have zero households reporting zero OOPs in both modules (as both of them had OOPs instruments in consumption modules administered frequently by visits throughout the year). On the other hand the surveys from Uganda show 30%+ of households reporting zero OOPs in both modules.

Ghana_1991 has the worst non-correspondant rates where 57.9% of households reported positive OOPs in the health or consumption module, and then reported zero on the other one. This is solely driven by the zero OOPs responses in the health module, which happened to be episodic, so in order for individuals to report OOPs they needed to report being recently ill. On the other hand, the non-concordance rate of Nigeria_2015 is completely explained by its consumption module, the OOPs instrument happened to be a single question asking households for their total OOPs.



```{r table_zero_concordance, echo=FALSE}

# Put it in percentage
percentage <- function(x){
  round(x*100,1)
  }

tab_zero_reps_per <- tab_zero_reps %>% 
  mutate_at(vars(matches("h")) , percentage) %>% 
  mutate(`Concordance Rate` = both_zero + posc_poch,
         `Non-Concordance Rate` = 100- (both_zero + posc_poch) ) %>% 
  rename(`Both Zero` = both_zero,
         `C>0, H=0` = posc_zeroh,
         `C=0, H>0` = zeroc_posh,
         `Both Positive` = posc_poch,
         Survey = survey)

  kable(tab_zero_reps_per, caption = "Concordance between OOPs by survey" ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
    footnote(general = "The columns represent the percentage of households in each category depending on their reported OOPs in the consumption and health module. The first column is the percentage of households in the survey that reported zero OOPS in both the health module and consumption module. The second column is the percentage of households reporting strictly positive OOPs in the consumption module, and zero OOPs in the health module. ")
```
# Mean OOPs by Module and socio-economic status

We have seen so far that OOPs from the health module tend to be higher than those from the consumption module. A possibility is that these results are driven by socio-economic status, where only households from richer or poorer households report more OOPs a module. We will use the quintiles of non health consumption as a sign of socio-economic status, excluding OOPs is important as total consumption is not a relevant social marker for households experiencing a health shock and had to go through high health care expenditures.

Figure 3 exhibits for each survey and consumption quintile the mean OOPs of the consumption module over those of the health module. We observe no clear pattern between socio-economic status and the differences in the reported OOPs between both modules.


```{r mean_consumption_quintile, echo=FALSE}

# Grab means by survey and module, and consumption quintile
tab_mean_cq <- df_bothm %>% 
  group_by(survey,module,consumption_quintile) %>% 
  summarise(mean_oops = mean(oops, na.rm = TRUE),
            median_oops = median(oops, na.rm=TRUE),
            obs = n())

# Grab OOPs of health module over OOPs of consumption module
tab_hc_cq <- tab_mean_cq %>% 
  group_by(survey,consumption_quintile) %>% 
  mutate(healthc_healthm = first(mean_oops) / last(mean_oops),
         mhealthc_mhealthm = first(median_oops) / last(median_oops),
         trim_mhealthc_mhealthm = if_else(mhealthc_mhealthm>2,2,mhealthc_mhealthm)) %>% 
  ungroup() %>% 
  select(survey, consumption_quintile, mhealthc_mhealthm, healthc_healthm, trim_mhealthc_mhealthm) %>% 
  distinct() %>% 
  mutate( order = seq_along(survey),
          survey2 = fct_reorder(survey, desc(order) )) %>% # Create order for surveys
  filter(consumption_quintile> -1) # eliminate those with NA


# Graph mean
tab_hc_cq %>% 
  ggplot(aes(x=survey2,y=healthc_healthm, 
             group = reorder(consumption_quintile, -consumption_quintile), 
             fill = consumption_quintile) ) +
  geom_bar(position='dodge', stat='identity') +
  coord_flip() +
  ggtitle("Mean OOPs from the consumption module over the health module") +
  labs(subtitle = "By consumption quintile") +
  xlab("") +
  ylab("Share") +
  geom_hline(yintercept=1, alpha=3, linetype = "dashed")


```

# Catastrophic health expenditures (CHEs)

Popular datasets like the World's Bank HEFPI, and the World Development Indicators include estimations for the percentage of households incurring into a level of OOPs that is too burdensome to bear. The most popular indicators are Catastrophic health expenditures, which consist on a ratio between the OOPs of a household and its total available resources. What is troubling, is that these popular datasets publish their estimates without stating if the OOPs were coming from the health or consumption module, and without awareness of what are the possible effects of these.

Previously, we saw that OOPs tended to be higher for the health module, now let's check how much this can potentially impact CHE meaures. We will use our sample of surveys with OOPs in a consumption and health module, and add these two estimates of OOPs to the non-health consumption of each household to come up with two different estimates of total consumption. Then, we will be able to calculate the most popular indicator of the CHE kind, 10% of total consumption, which is simply an binary variable of whether a household's OOPs exceed 10% of their total consumption.













